The pollination trade-off

### 2019-03-17 paper/manuscript.Rmd
Add full initials to my name
Complete edits to discussion

---
title: "The pollination trade-off"
author: "Fernando Cagua^1^, Hugo J. Marrero^2^, Jason M. Tylianakis^1^, Daniel B. Stouffer^1^"
bibliography: phd-bibliography/pollen-competition.bib
csl: ecology-letters.csl
output:
  bookdown::pdf_document2:
    # base_format: rticles::peerj_article
    keep_tex: yes
    number_sections: false
toc: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{lineno}
  - \usepackage{xr}
  - \usepackage[utf8]{inputenc}
  - \newcommand{\R}[1]{\label{#1}\linelabel{#1}}
  - \newcommand{\lr}[1]{page~\pageref{#1}, line~\lineref{#1}}
  - \externaldocument[S-]{supp-info}
---

@@

Both of the tests we used to identify the effect of animal-mediated pollination suggested the same outcome: that net facilitation is only experienced by a fraction of species in a community. 
A recent study also suggested a wide range of outcomes in a multi-species context [@tur_evaluating_2016]. 
However, the factors explaining why some species experience more competition than others were not understood. 
Here we show that the prevalence of neutral and competitive effects of animal-mediated pollination can be at least partially explained by the trade-offs that minimising competition can involve.  
These trade-offs were most evident in precisely the factors that most strongly modulate pollen deposition. 
Specifically, we found that the factors related to the relative dominance of the species (in terms of pollen and visits, or abundance) also tended to be the most important predictors of pollen deposition. 
At the same time, receiving high visitation increases the absolute levels of conspecific pollen deposition but increasses heterospeicific deposition to a greater extent even when the visitors are likely carry a high proportion of conspecific pollen. 
Similarly, being abundant reduces the amount of heterospecific pollen deposited but at the expense of a noticeable reduction in the amount of conspecific pollen. 
Multiple studies recognise the importance of visitation and abundance to competition for pollination, however, some suggest a positive effect on the pollination service while some others a negative one [@feldman_when_2004; @munoz_presence_2008; @morales_interspecific_2008]. 
Our results corroborate the importance of these two factors, and show that the contrsating results obtained by earlier studies could be explained by the trade-offs involved.

Simultaneously, because visitation, pollen production and abundance are correlated [@sargent_role_2006], the trade-offs also suggest that plants must reach a fine balance between increasing conspecific pollen and reduce heterospecific pollen deposition to minimise competition for pollination. 
The relatively broad distribution of coefficients in our models might be explained by the large variation of the extent to which plant species are affected by pollen deposition. 
Indeed, some plant species experience a large reduction in seed set due to heterospecific pollen deposition, while others are relatively insensitive to heterospecific pollen but greatly affected by low levels of conspecific pollen deposition [@campbell_mechanism_1985].

The two other strategies we analysed involved a smaller trade-off than the abundance or visit efficacy. 
Specifically, having high functional originality relative to other species in the community or sharing few pollinators was found to be beneficial to both the quantity and quality of pollination.
Interestingly, these two strategies also had lower explanatory power in the models of pollen deposition. 
This does not necessarily imply that these variables are unimportant to pollen deposition. 
Instead, it is possible that the processes that underlay these strategies might be more complex than what our metrics are able to capture. 
Particularly so for conspecific pollen deposition, which was harder to predict than heterospecific pollen. 
For instance, having distinct traits, for example through phenology differentiation, might have a positive impact on the pollination service only to the extent to which previous floral communities have promoted high pollinator densities (ref). 
In addition, some plants have been shown possess mechanisms that prevent heterospecific pollen deposition even when shared visitors carry pollen loads from diverse species. 
Alternatively, other factors that we were unable to measure, like pollinator behaviour and densities [@ye_competition_2014; @flanagan_effects_2011; @melissas_2019?] or the spatial context [@cariveau_spatially_2009], can also shape the way sharing pollinators affect the pollination service.

\R{strategies}Regardless of the specific factors involved it is clear, however, that the community context has a very important effect on the way plant species respond to competition for pollination. 
First, in the models of pollen deposition, the random structure that allowed plants of the same species to have different intercepts across communities was much better supported by the data than the structure that fixed the intercept for a plant species irrespective of the community. 
Second, in a large majority of cases, the strategies that a species uses across communities are not more similar to each other than what would be expected at random. 
Lit? Context?

>>>>

---
title: "The pollination trade-off"
author: "E. Fernando Cagua^1^, Hugo J. Marrero^2^, Jason M. Tylianakis^1^, Daniel B. Stouffer^1^"
bibliography: phd-bibliography/pollen-competition.bib
csl: ecology-letters.csl
output:
  bookdown::pdf_document2:
    # base_format: rticles::peerj_article
    keep_tex: yes
    number_sections: false
toc: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{lineno}
  - \usepackage{xr}
  - \usepackage[utf8]{inputenc}
  - \newcommand{\R}[1]{\label{#1}\linelabel{#1}}
  - \newcommand{\lr}[1]{page~\pageref{#1}, line~\lineref{#1}}
  - \externaldocument[S-]{supp-info}
---

@@

Both of the tests we used to identify the effect of animal-mediated pollination suggested the same outcome: that net facilitation is only experienced by a fraction of species in a community. 
A recent study also suggested a wide range of outcomes in a multi-species context [@tur_evaluating_2016]. 
However, the factors explaining why some species experience more competition than others were not understood. 
Here we show that the prevalence of neutral and competitive effects of animal-mediated pollination can be at least partially explained by the trade-offs that minimising competition can involve.

These trade-offs were most evident in precisely the factors that most strongly modulate pollen deposition. 
Specifically, we found that the factors related to the relative dominance of the species (in terms of pollen and visits, or abundance) also tended to be the most important predictors of pollen deposition. 
At the same time, receiving high visitation increases the absolute levels of conspecific pollen deposition but increasses heterospeicific deposition to a greater extent---even when the visitors are likely carry a high proportion of conspecific pollen. 
Similarly, being abundant reduces the amount of heterospecific pollen deposited but at the expense of a noticeable reduction in the amount of conspecific pollen. 
Multiple studies recognise the importance of visitation and abundance to competition for pollination [@feldman_when_2004; @munoz_presence_2008; @morales_interspecific_2008]. 
Our results strongly corroborate the importance of these two factors but also suggest that [because visitation, pollen production and abundance are correlated; @sargent_role_2006], plants must reach a fine balance between increasing conspecific pollen and reduce heterospecific pollen deposition to minimise competition for pollination.

Further evidence of the intricacies of balancing costs and benefits can be found when we examine the other two factors analysed in our models. 
We show that being a specialist and sharing no pollinators might be preferable in the short term (as it is generally detrimental to pollen deposition). 
However, over long periods of time there could be a risk associated to a specialist plant having no pollinators. 
Hence, is likely that, to ensure long term survival, plants also need to balance this risk with the costs of sharing pollinators. 
One possible solution is to share pollinators *and* have original traits---as we show that trait originallity is generally benefitial to pollen deposition and it’s commonly thought that species that are further from others in trait space suffer reduced competition. 
Yet again, there are two possible caveats to this strategy that highlight the interrelatedness of the ecological factors. 
First, in a mutualism context, it is also possible that trait originality could come at the cost of being less ‘apparent’ to pollinators. 
Second, the usuallly negative, relationship between originality and generalism has been shown depend on the plants abundance [@coux_linking_2016], with generalist species being able to also have original traits only when they are abundant enough to provide a valuable reward to make it worth to pollinators.

How precisely plants balance the costs and benefits in the short and the long term likely depends on the species sensitivity to different risks but also, as we show here, on the community it belongs to. 
For instance, a plant species that experience a large reduction in seed set due to heterospecific pollen deposition might be better off by sharing few pollinators, while sharing pollinators makes more sense for those mainly affected by low levels of conspecific pollen deposition [@campbell_mechanism_1985].
Nevertheless, the diversity of the facilitation/competition outcomes across communities we observed and the relevance of random effects in our models suggests that the community context has a very important effect on the way plant species respond to competition for pollination beyond species identity. 
Perhaps overgeneralising, some plant families (Asteraceae for example) are often generalists in their communities, others (Orchidaceae) are known to be more commonly specialised. 
While it is possible, from an evolutionary perspective, that selection for a particular strategy might say something about the community in which a species has typically inhabited during its evolutionary history, we show that most plant species are still flexible enough to adopt markedly different niches in different communities. 
From a more applied perspective, flowering plants are sometimes introduced to attract pollinators on other nearby plants. 
Our results highlight the difficulties involved with predicting wether the introduced plant species will facilitate or compete with neighbours. 
This is so because the outcome might depend not only on the particular niche each species occupies in a particular community, but also on other factors that we were unable to measure [e.g. pollinator behaviour and densities or spatial context; @ye_competition_2014; @flanagan_effects_2011; @cariveau_spatially_2009].
However, our results also suggest that introduced species are more likely to be benefitial if they have differing traits to local plants. 
<!-- cite melissa's paper if available before publication -->
